---
title: "Least Squares Best Fit Line"
author: "Michael Ghattas"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Find Minimum Square Error Line

View the data. These are from a *Pets for Life* data set provided to the author. 

```{r}
dat<-read.csv("animal_stats_compact.csv")

(g<-ggplot(dat, aes(x=cat,y=dog))+geom_point())
```

Calculate the slope and intercept according to the formula.

* Recall that the slope in the $y=mx+b$ model according to the least squares criterion is $\frac{\frac{1}{n}\sum{x_iy_i-\bar{x}\bar{y}}}{\frac{1}{n}\sum{x_i^2-\bar{x}^2}}$ and the intercept is $\bar{y}-m\bar{x}$.*

```{r}
m<-(mean(dat$cat*dat$dog)-mean(dat$cat)*mean(dat$dog))/(mean(dat$cat^2)-mean(dat$cat)^2)
b<-mean(dat$dog-m*dat$cat)

g<-g+geom_abline(slope=m,intercept = b)
g<-g+labs(title="Least Squares Line",x="cat count",y="dog count")
g

```

Or use minimization directly. (This is just for illustration. It is not a recommended way to do this calculation.)

```{r}
sq_error<-function(x){
  return(sum((dat$dog-x[2]*dat$cat-x[1])^2))
}
nlm(sq_error,p=c(0,1))$estimate

```

There is a built-in function for this. The formula $dog=m\left(cat\right)+b$ is represented by "dog~cat".

```{r}
lm(dog~cat,data=dat)$coefficients
```

# Practice

* In the code block below, please calculate the estimated slope and intercept from the formulas $\frac{\frac{1}{n}\sum{x_iy_i-\bar{x}\bar{y}}}{\frac{1}{n}\sum{x_i^2-\bar{x}^2}}$ and \ $\bar{y}-m\bar{x}$, respectively, but this time modeling "dog" as the x-variable and "cat" as the y-variable.Output your values for "m" and "b". Check your formulas using the appropriate call to "lm".*

```{r}
sq_error<-function(x){
  return(sum((dat$cat - x[2] * dat$dog - x[1])^2))
}
nlm(sq_error,p=c(0,1))$estimate

lm(cat~dog, data = dat)$coefficients
```

