---
title: "Inference Verification"
author: "Michael Ghattas"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(HistData)
```

# Inference Formulas

One purpose of the following code is to connect the inference formulas from the theory of linear regression to the output of the model-fitting code. A second purpose is to provide visualizations for some of the values in the inference formulas.

## Data and Model

Remain with the model of Wealth as a linear function of Commerce in central France around 1830.

```{r}
data("Guerry")
dat<-Guerry

dat.c<-filter(dat,Region=="C")
m.c.w<-lm(Wealth~Commerce,data=dat.c)
summary(m.c.w)

```


## Some model sums of squares

```{r}
# SSY
(SSY<-sum((dat.c$Wealth-mean(dat.c$Wealth))^2))

# SSE
(SSE<-sum((dat.c$Wealth-m.c.w$fitted)^2))

# SSR
(SSR<-sum((m.c.w$fitted-mean(dat.c$Wealth))^2))

SSY
SSE+SSR
```

$R^2=\left(SSY-SSE\right)/SSY=SSR/SSY$, the percent of the variation in the response variable accounted for by the fitted values. Recall that $R^2$ equals the square of the correlation of the response variable and the explanatory variable.

# Practice 1

*In the code block below, please compute $R^2=SSR/SSY$. Also, please compute the square of the correlation of "dat.c$Wealth" and "dat.c$Commerce". Confirm that both results equal the value of $R^2$ given in the summary of the model "(m.c.w)".*

```{r}
(SSR/SSY)

cor(dat.c$Wealth,dat.c$Commerce)
cor(dat.c$Wealth,dat.c$Commerce)^2

summary(m.c.w)$r.squared
```

## p-value of regression

*The p-value of the regression can be calculated from the statistic $\frac{\frac{SSY-SSE}{\left(n-1\right)-\left(n-2\right)}}{\frac{SSE}{n-2}}$. Under the null hypothesis that $m=0$, this has an F-distribution with numerator degrees of freedom equal to $\left(n-1\right)-\left(n-2\right)=1$ and denominator degrees of freedom equal to $n-2$.*

```{r}
n<-nrow(dat.c)
(s2<-SSE/(n-2))
(f.stat<-SSR/s2)
(pf(f.stat,1,n-2,lower.tail=FALSE))
(pf((SSY-SSE)/((n-1)-(n-2))/(SSE/(n-2)),1,n-2,lower.tail=FALSE ))

summary(m.c.w)
```


## Standard error of slope

# Practice 2

*Compute the standard error of the slope. Confirm the p-value for the slope. Recall that $s^2\frac{1}{\sum_{i=1}^n\left(X_i-\bar{X}\right)^2}$ is a formula for the standard error of the slope. This is equivalent to $s^2\frac{1}{\sum_{i=1}^nX_i^2-n\bar{X}^2}$. The value $s^2$ is stored in the variable "s2" above. Confirm that the standard error computed this way equals the standard error from the summary of "m.c.w".*

```{r}
sqrt(s2/sum((dat.c$Commerce-mean(dat.c$Commerce))^2))
summary(m.c.w)$coefficients[2,2]


# Calculate the p-value of the slope from the definition.
2*pt(-abs(summary(m.c.w)$coefficients[2,1]/summary(m.c.w)$coefficients[2,2]),df=n-2)

# Verify that this is the p-value from the summary.
summary(m.c.w)$coefficients[2,4] 



```

## Standard error of intercept

*Recall that $$s^2\left(\frac{1}{n}+\frac{\bar{X}^2}{\sum_{i=1}^n\left(X_i-\bar{X}\right)^2}\right)$$ is a formula for the standard error of the intercept. This is equivalent to $s^2\left(\frac{1}{n}+\frac{\bar{X}^2}{\sum_{i=1}^nX_i^2-n\bar{X}^2}\right)$.*

```{r}
sqrt(s2*(1/n+mean(dat.c$Commerce)^2/sum((dat.c$Commerce-mean(dat.c$Commerce))^2)))
summary(m.c.w)$coefficients[1,2] # Verify this is the Std. Error from the summary.


# Calculate the p-value of the intercept from the definition.
2*pt(-abs(summary(m.c.w)$coefficients[1,1]/summary(m.c.w)$coefficients[1,2]),df=n-2)

# Verify that this is the p-value from the summary.
summary(m.c.w)$coefficients[1,4] 
```


## Predicted Y Random Variables

In the model that $Y$ is a linear function of $X$ plus iid Normal errors $\varepsilon$, the maximum likelihood slope and intercept are random variables, functions of $\varepsilon$. Thus the predicted value of $\hat Y_h$ at $X_h$ is a random variable distributed as $\hat{Y}_{h}\sim Normal\left(mX_{h}+b,\sigma^{2}\left(\frac{1}{n}+\frac{\left(X_{h}-\bar{X}\right)^{2}}{\sum\left(X_{i}-\bar{X}\right)^{2}}\right)\right)$. The confidence interval based on this uses the corresponding Student's t distribution with $n-2$ degrees of freedom, replacing $\sigma^2$ by $s^2$, that is $\frac{SSE}{n-2}$.

```{r}

g<-ggplot(data=dat.c,aes(x=Commerce,y=Wealth))+geom_point()
g<-g+geom_abline(slope=m.c.w$coefficients[2],intercept=m.c.w$coefficients[1])
g

```

### Look at 95% conf. interval for Y.pred

```{r}
x.new<-seq(min(dat.c$Commerce),max(dat.c$Commerce),length.out=50)
temp<-data.frame(Commerce=x.new)
y.pred<-predict(m.c.w,newdata=temp)
```

### function to calculate variance for the mean of a new observation

```{r}
s2.new<-function(newx,x,s2){
	n<-length(x)
	return(s2*(1/n+(newx-mean(x))^2/sum((x-mean(x))^2)))
}
```

### Generate vector of upper and lower bounds for the means 

Use the 95% confidence interval. Plot the results.

```{r}
s2s<-s2.new(x.new,dat.c$Commerce,s2) # variance of the mean of an observation at x.new
a<-qt(.975,n-2) # .975 quantile for Student's t with n-2 degrees of freedom
upper<-y.pred+a*sqrt(s2s) # upper bound of 95% CI on y.pred
lower<-y.pred-a*sqrt(s2s) # lower bound of 95% CI on y.pred

dat.new<-data.frame(x=x.new,lower=lower,upper=upper)
g<-ggplot(dat.c,aes(x=Commerce,y=Wealth))+geom_point()+geom_smooth(method="lm")+
  geom_line(data=dat.new,aes(x=x,y=lower),color="green")+
  geom_line(data=dat.new,aes(x=x,y=upper),color="green")
g

```

The distribution of a new $\hat Y_h new$ is given by $\hat{Y}_{h}new\sim Normal\left(mX_{h}+b,\sigma^{2}\left(\frac{n+1}{n}+\frac{\left(X_{h}-\bar{X}\right)^{2}}{\sum\left(X_{i}-\bar{X}\right)^{2}}\right)\right)$. Again, the confidence intervals corresponding to this use a Student's t distribution with $n-2$ degrees of freedom, replacing $\sigma^2$ by $s^2$.

# Practice 3

*Please supply the upper and lower bounds on the 95% confidence intervals for new observations as the indicated columns in dat.new. Run the plotting commands to view these bounds.*

```{r}
# y.pred.new, with error
dat.new$upper.w.error<-y.pred+a*sqrt(s2s+s2)
dat.new$lower.w.error<-y.pred-a*sqrt(s2s+s2)
g<-g+geom_line(data=dat.new,aes(x=x,y=lower.w.error),color="orange")+
  geom_line(data=dat.new,aes(x=x,y=upper.w.error),color="orange")
g
```