---
title: "dataframes"
author: "Michael Ghattas"
output:
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Part 1 Examples

## Read in data

General text files may be read in using *read.table* using appropriate arguments. CSV format, an "industry standard" typically uses *read.csv*

```{r}
dat<-read.csv("worms.csv")
```

## Inspecting the data

You can view dataframes by double clicking in RStudio. There are some standard summary tools.

```{r}
dat
dim(dat)
names(dat)
str(dat)
summary(dat)
head(dat)
dat[2,4]
```

## Select columns

Leaving the row index blank and providing column indices allows selection of a subset of the columns.

```{r}
dat[,3]
dat$Vegetation
dat[,2:4]
```

## Select rows

Leaving the column index blank and providing a vector of indices or boolean values allows selection of a subset of the rows.

```{r}
dat[1:4,]
damp<-dat[dat$Damp,]
summary(damp)
flat<-dat[dat$Slope<4.5,]
summary(flat)
```

# Practice Problem for Part 1

*Please supply code to create a new data frame "grass" the restricts "dat" to observations in which the "Vegetation" variable equals "Grassland". Output the values of the variable "Damp" for this data frame.*

```{r}
# Create the 'grass' data frame
grass <- subset(dat, Vegetation == "Grassland")

# Output the values of the 'Damp' variable for the 'grass' data frame
damp_values <- grass$Damp

# Display the 'grass' data frame and 'Damp' values
grass
damp_values
```

# Part 2 Example

## Find Maximum Likelihood, binomial

Maximize $p^{30}\left(1-p\right)^{20}$. The "optimize" function works for single variable functions.

```{r}
# function to calculate and return x^30*(1-x)^20, the function to be optimized.
f<-function(x){
  return (x^30*(1-x)^20)
}
(ml_ext<-optimize(f,c(0,1),maximum=TRUE)) # Look for the value that maximizes the function f 
                                          # in the range [0,1], the range of valid probabilities.
30/50 # theoretical value =30/(30+20)
f(.6) # check
```

# Practice Problem for Part 2

*Please revise the code above to calculate the value of $p$ that maximizes* $p^{15}\left(1-p\right)^{85}$. *Check the result by computing the relevant ratio. Calculate the maximum value of the function.*

```{r}
# Function to calculate and return p^15*(1-p)^85, the function to be optimized.
f <- function(p) {
  return (p^15 * (1-p)^85)
}

# Look for the value that maximizes the function f in the range [0,1].
ml_ext <- optimize(f, c(0, 1), maximum = TRUE)

# Theoretical value
theoretical_value = (15 / (15 + 85))

# Check the value at the theoretical maximum
f_theoretical <- f(theoretical_value)

# Calculate the maximum value of the function
max_value <- ml_ext$objective

# Output the results
ml_ext
theoretical_value
f_theoretical
max_value
```
