---
title: "Midterm"
author: "Michael Ghattas"
output:
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Instructions

Please work these problems on your own. You may use web searches and
refer to your notes, lectures, and past problem sets, but not
interactive methods such as asking others online or in person.

# Question 1

## Context for Question 1

A medical researcher investigates whether systolic blood pressure
readings depend on the arm, left or right, from which an individual's
blood pressure is measured. The researcher takes readings from both arms
of 100 individuals and records $L$ if the measurement from the left arm
is higher and $R$ if the measurement from the right arm is higher.
Consider the model that the number of $L$s follows a binomial
distribution with the size equal to $100$ and the probability of success
equal to $0.5$. That is, the probability that $k$ individuals have
measurements that result in $L$ is
$f(k)=\left(\begin{array}{c}100\\k\end{array}\right)(0.5)^{k}(1-0.5)^{100-k}=\left(\begin{array}{c}100\\k\end{array}\right)(0.5)^{100}$.
Denote this probability space by $binomial(100,0.5)$.

## Question 1.1

(3 points)

What is the probability of the event that the number of $L$s is equal to
$50$ under the $binomial(100,0.5)$ model?

**Your code and answer here:** For this specific problem, we have
$n = 100$, $k = 50$, and $p = 0.5$:
$P(X = 50) = \binom{100}{50} (0.5)^{50} (0.5)^{50} = \binom{100}{50} (0.5)^{100}$.
The binomial coefficient $\binom{100}{50}$ can be calculated as:
$\binom{100}{50} = \frac{100!}{50! \cdot 50!}$ Thus, the probability is:
$P(X = 50) = \frac{100!}{50! \cdot 50!} \cdot (0.5)^{100}$. Using R, we
find:

```{r}
# Calculate the probability of getting exactly 50 Ls out of 100
prob_50_Ls <- dbinom(50, size = 100, prob = 0.5)
prob_50_Ls
```

## Question 1.2

(3 points)

What is the probability of the event that the number of $L$s is less
than or equal to $40$ under the $binomial(100,0.5)$ model?

**Your code and answer here:** To find the probability that the number
of $L$s is less than or equal to 40, we need to sum the probabilities
for $k = 0$ to $k = 40$:
$P(X \leq 40) = \sum_{k=0}^{40} \binom{100}{k} (0.5)^k (0.5)^{100-k} = \sum_{k=0}^{40} \binom{100}{k} (0.5)^{100}$.
Using R, we can calculate this cumulative probability:

```{r}
# Calculate the cumulative probability of getting 40 or fewer Ls out of 100
prob_40_or_less_Ls <- pbinom(40, size = 100, prob = 0.5)
prob_40_or_less_Ls
```

## Question 1.3

(3 points)

If the researcher treats $binomial(100,0.5)$ as the null distribution
for the null hypothesis that neither arm consistently produces larger
measurements, is an observed count of $L$s equal to $60$ strong evidence
against the null hypothesis? Please explain.

**Your code and answer here:** We can use the cumulative distribution
function (CDF) of the binomial distribution to find the probability of
observing 60 or more $L$s: $P(X \geq 60) = 1 - P(X \leq 59)$. Where
$P(X \leq 59)$ is the cumulative probability up to 59. Using R, we can
calculate this cumulative probability and the resulting $p$-value. A
small $p$-value (typically less than 0.05) indicates strong evidence
against the null hypothesis.

```{r}
# Calculate the cumulative probability of getting 59 or fewer Ls out of 100
prob_59_or_less_Ls <- pbinom(59, size = 100, prob = 0.5)
# Calculate the probability of getting 60 or more Ls out of 100
p_value <- 1 - prob_59_or_less_Ls
p_value
```

Since the calculated $p-value$ is less than 0.05, it indicates that an
observed count of $L$'s equal to 60 is strong evidence against the null
hypothesis that neither arm consistently produces larger measurements.

## Question 1.4

(3 points)

If the researcher treats $binomial(100,0.5)$ as the null distribution
for the null hypothesis that neither arm consistently produces larger
measurements, what would be an unusually large or small observation
under the null hypothesis? Please explain your answer and how you
quantify unusual?

**Your code and answer here:** We need to find the critical values $k$
such that:
$P(X \leq k_{\text{low}}) \leq 0.025 \quad \text{and} \quad P(X \geq k_{\text{high}}) \leq 0.025$.
Using the cumulative distribution function (CDF) for the binomial
distribution, we can find these critical values. In R, this can be
achieved using the `qbinom` function.

```{r}
# Define the significance level
alpha <- 0.05
# Find the critical value for the lower tail (2.5%)
k_low <- qbinom(alpha / 2, size = 100, prob = 0.5)
# Find the critical value for the upper tail (97.5%)
k_high <- qbinom(1 - alpha / 2, size = 100, prob = 0.5)
list(k_low = k_low, k_high = k_high)
```

Therefore, an observation of 40 or fewer $L$'s would be unusually small,
and an observation of 60 or more $L$'s would be unusually large under
the null hypothesis. This quantification is based on the typical $5%$
significance level, meaning that such observations would occur by chance
less than $5%$ of the time under the null hypothesis.

# Question 2

Consider the experiment of tossing two fair 6-sided dice.

## Question 2.1

(3 points)

Please define a reasonable sample space and event space to model this
experiment.

**Your answer here:** The sample space $S$ for tossing two fair 6-sided
dice consists of all possible ordered pairs $(d_1, d_2)$ where $d_1$ and
$d_2$ represent the outcomes on the first and second die, respectively.
Since each die has 6 faces, there are a total of $6 \times 6 = 36$
possible outcomes:
$S = \{ (d_1, d_2) \mid d_1, d_2 \in \{1, 2, 3, 4, 5, 6\} \}$. An event
$E$ is a subset of the sample space $S$. Here are all possible outcomes:
$S = \{(1,1), (1,2), (1,3), (1,4), (1,5), (1,6), (2,1), (2,2), (2,3), (2,4), (2,5), (2,6), \\(3,1), (3,2), (3,3), (3,4), (3,5), (3,6), (4,1), (4,2), (4,3), (4,4), (4,5), (4,6), \\ (5,1), (5,2), (5,3), (5,4), (5,5), (5,6), (6,1), (6,2), (6,3), (6,4), (6,5), (6,6)\}$

## Question 2.2

(3 points)

What is the probability of an outcome $\left(a,b\right)$ where a is the
outcome of the first toss of the die, and b is the outcome of the second
toss of the die?

**Your answer here:** Since each die is fair and has 6 faces, each face
has an equal probability of appearing. The probability of any specific
outcome on one die is $\frac{1}{6}$. When tossing two fair 6-sided dice,
the outcome of each die is independent of the other. Therefore, the
probability of any specific pair $(a, b)$ occurring, where $a$ is the
outcome of the first die and $b$ is the outcome of the second die, is
the product of the probabilities of each individual outcome.
$P((a, b)) = P(a) \times P(b)$. Given that $P(a) = \frac{1}{6}$ and
$P(b) = \frac{1}{6}$, we have:
$P((a, b)) = \left( \frac{1}{6} \right) \times \left( \frac{1}{6} \right) = \frac{1}{36}$.
Therefore, the probability of any specific outcome $(a, b)$ where $a$ is
the outcome of the first toss and $b$ is the outcome of the second toss
is: $P((a, b)) = \frac{1}{36}$.

## Question 2.3

(3 points)

What is the probability of the event
$E = \{\left(a,b\right)|a\geq 3b\}$?

**Your answer here:** We need to determine the pairs $(a, b)$ where $a$
is at least three times $b$. Since both $a$ and $b$ can take values from
1 to 6 (because the dice are 6-sided), we identify the pairs that
satisfy $a \geq 3b$: 1. $b = 1: \{ (3,1), (4,1), (5,1), (6,1) \}$ 2.
$b = 2: \{ (6,2) \}$ For $b \geq 3$, there are no possible values of $a$
that satisfy $a \geq 3b$ because $a$ can only go up to 6. Thus, the
event $E$ consists of the following pairs:
$E = \{ (3,1), (4,1), (5,1), (6,1), (6,2) \}$. There are 5 such pairs.
The total number of possible outcomes when tossing two fair 6-sided dice
is 36. The probability of the event $E$ is the number of favorable
outcomes divided by the total number of possible outcomes:
$P(E) = \frac{|E|}{|S|} = \frac{5}{36}$. Therefore, the probability of
the event $E$ is: $P(E) = \frac{5}{36}$.

## Question 2.4

(3 points)

What is the probability of the event $B = \{\left(a,1\right)|b=1\}$ that
the second toss $b$ is has a value of 1?

**Your answer here:** To determine the probability of the event $B$
where the second die shows a 1, we need to identify all the pairs
$(a,1)$ where $a$ can be any value from 1 to 6, and $b$ is fixed at 1.
The possible pairs are:
$B = \{ (1,1), (2,1), (3,1), (4,1), (5,1), (6,1) \}$. There are 6 such
pairs. The total number of possible outcomes when tossing two fair
6-sided dice is 36. The probability of the event $B$ is the number of
favorable outcomes divided by the total number of possible outcomes:
$P(B) = \frac{|B|}{|S|} = \frac{6}{36} = \frac{1}{6}$. Therefore, the
probability of the event $B$ is: $P(B) = \frac{1}{6}$.

## Question 2.5

(3 points)

Are the events $B = \{\left(a,1\right)|b=1\}$ and
$E = \{\left(a,b\right)|a\geq 3b\}$ independent?

**Your answer here:** To determine if the events $B$ and $E$ are
independent, we need to check if: $P(B \cap E) = P(B) \cdot P(E)$.
First, recall the definitions and probabilities of the events: 1. Event
$B$: The second die shows a 1,
$B = \{ (1,1), (2,1), (3,1), (4,1), (5,1), (6,1) \}$,
$P(B) = \frac{6}{36} = \frac{1}{6}$. 2. Event $E$: The first die is at
least three times the second die,
$E = \{ (3,1), (4,1), (5,1), (6,1), (6,2) \}$, $P(E) = \frac{5}{36}$.
Now, consider the intersection
$B \cap E = \{ (3,1), (4,1), (5,1), (6,1) \}$. There are 4 outcomes in
$B \cap E$. The probability of the intersection is:
$P(B \cap E) = \frac{|B \cap E|}{|S|} = \frac{4}{36} = \frac{1}{9}$.
Now, we compare $P(B \cap E)$ with
$P(B) \cdot P(E) = \frac{1}{6} \cdot \frac{5}{36} = \frac{5}{216}$. We
need to see if: $P(B \cap E) = P(B) \cdot P(E)$,
$\frac{1}{9} = \frac{5}{216}$. Clearly,
$\frac{1}{9} \neq \frac{5}{216}$. Therefore, the events $B$ and $E$ are
not independent because $P(B \cap E) \neq P(B) \cdot P(E)$.

# Question 3

Consider a continuous random variable $X$ with the probability density
function defined by $f(x)=4x^3$ for $x\in[0,1]$ and $f(x)=0$ otherwise.

## Question 3.1

(3 points)

What is cumulative distribution of $X$?

**Your answer here:** Given the PDF $f(x) = 4x^3$ for $x \in [0,1]$ and
$f(x) = 0$ otherwise, we can find the CDF $F(x)$ for different ranges of
$x$. 1. For $(x < 0): F(x) = 0$ 2. For
$(0 \leq x \leq 1): F(x) = \int_{0}^{x} 4t^3 \, dt$ 3. For $x > 1$:
$F(x) = 1$ (We calculate the integral:
$F(x) = 4 \int_{0}^{x} t^3 \, dt = 4 \left[ \frac{t^4}{4} \right]_{0}^{x} = 4 \left( \frac{x^4}{4} - 0 \right) = x^4$).
Thus, the cumulative distribution function $F(x)$ of $X$ is: $$
F(x) = 
\begin{cases} 
0 & \text{if } x < 0 \\
x^4 & \text{if } 0 \leq x \leq 1 \\
1 & \text{if } x > 1 
\end{cases}
$$

## Question 3.2

(4 points)

What is the expected value of $X$, $E[X]$?

**Your answer here:** Given the PDF $f(x) = 4x^3$ for $x \in [0,1]$ and
$f(x) = 0$ otherwise, we can find the expected value $E[X]$ by
integrating over the range where $f(x)$ is non-zero:
$E[X] = \int_{0}^{1} x \cdot 4x^3 \, dx$. Simplify the integrand:
$E[X] = \int_{0}^{1} 4x^4 \, dx$. Now, calculate the integral:
$E[X] = 4 \int_{0}^{1} x^4 \, dx = 4 \left[ \frac{x^5}{5} \right]_{0}^{1} = 4 \left( \frac{1^5}{5} - \frac{0^5}{5} \right) = 4 \cdot \frac{1}{5} = \frac{4}{5}$.
Therefore, the expected value of $X$ is: $E[X] = \frac{4}{5}$.

## Question 3.3

(4 points)

What is the variance of $X$, $Var[X]$?

**Your answer here:** Given the PDF $f(x) = 4x^3$ for $x \in [0,1]$ and
$f(x) = 0$ otherwise, we can find $E[X^2]$ by integrating over the range
where $f(x)$ is non-zero: $E[X^2] = \int_{0}^{1} x^2 \cdot 4x^3 \, dx$.
Simplify the integrand: $E[X^2] = \int_{0}^{1} 4x^5 \, dx$. Now,
calculate the integral:
$E[X^2] = 4 \int_{0}^{1} x^5 \, dx = 4 \left[ \frac{x^6}{6} \right]_{0}^{1} = 4 \left( \frac{1^6}{6} - \frac{0^6}{6} \right) = 4 \cdot \frac{1}{6} = \frac{4}{6} = \frac{2}{3}$.
We already found the expected value $E[X] = \frac{4}{5}$. Now, we can
find the variance
$\text{Var}[X] = E[X^2] - (E[X])^2 = \frac{2}{3} - \left( \frac{4}{5} \right)^2 = \frac{2}{3} - \frac{16}{25} = \frac{2}{75}$.
Therefore, the variance of $X$ is: $\text{Var}[X] = \frac{2}{75}$.

# Question 4

Consider the hypothetical scenario in which a pharmaceutical company is
determining effect of an experimental drug on a rare disease.
Individuals in this study (800 individuals) are randomly assigned to the
treatment (in which the individual receives the drug) and (1000 are
assigned) to the placebo. The matrix m below represents the data in this
hypothetical scenario. Consider the null hypothesis that individuals who
receive the placebo and individuals who receive the treatment have the
same susceptibility to the disease. One possible probability model for
this is that each case in the pooled `Placebo` and `Treatment` cases can
be viewed as being assigned to the `Treatment` group with probability
equal to the proportion of the `Treatment` population in the pooled
`Placebo` and `Treatment` population. In the questions below we will
assess the claim that the drug has no effect.

```{r TreatmentDrug}
d<-rbind(c(800,3), c(1000,13))
m<-as.matrix(d)
colnames(m)= c("Population","Disease")
rownames(m) = c("Treatment", "Placebo")
m
```

## Question 4.1

(3 points)

Using the same null model described above, please calculate the
probability that the disease cases in the treatment group under the null
model is less than or equal to `m$Disease[1]` directly. Recall that the
function pbinom(x,size,prob) returns the probability of the event that
the number of successes is in the set $\{0,1,...x\}$.

**Your code and answer here:**

```{r BinomialProb}
# Define the parameters
size = 16
prob = (800 / 1800)
x = 3

# Calculate the probability using the binomial distribution function
probability <- pbinom(x, size = size, prob = prob)
probability
```

## Question 4.2

(3 points)

Use the function `rbinom(n,size,prop)` to simulate the number of
individuals with the disease in the treatment group under the null model
and null hypothesis. Use the function to generate 10,000 random numbers
representing the instances of the disease in the treatment group under
the null model and null hypothesis, and assign these 10,000 random
numbers to `sim_counts` in the code chunk below. Discuss each parameter
used in the rbinomial function x, size, and prob.

**Your code and answer here:**

```{r sim}
set.seed(1)
n = 10000
size = 16
prob = (800 / 1800)
sim_counts <- rbinom(n, size = size, prob = prob)
```

## Question 4.3

(3 points)

Use ggplot to display a histogram of `sim_counts` with the appropriate
number of bins. On the same graph with the histogram display the number
of actual observed counts of the disease in the treatment group.

**Your code and answer here:**

```{r hist}
# Load necessary library
library(ggplot2)

# Create a data frame for ggplot
sim_data <- data.frame(sim_counts = sim_counts)

# Create the histogram
ggplot(sim_data, aes(x = sim_counts)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = 3), color = "red", linetype = "dashed", size = 1) +
  labs(title = "Histogram of Simulated Disease Counts in Treatment Group",
       x = "Number of Disease Cases",
       y = "Frequency") +
  theme_minimal()
```

## Question 4.4

(3 points)

What would be an usually small number of observed counts of the disease
in the treatment group under the null model/hypothesis? Quantify what
you mean by unusual.

**Your code and answer here:**

```{r unusual}
# Calculate the 5th percentile of the simulated counts
lower_bound <- quantile(sim_counts, 0.05)
lower_bound
```

An unusually small number of observed counts of the disease in the
treatment group under the null model/hypothesis would be any number less
than or equal to 4. This is because the 5th percentile of the
distribution of simulated counts is 4, meaning that only 5% of the
simulations resulted in 4 or fewer cases. Therefore, we can consider
observing 4 or fewer cases as unusual under the null hypothesis.

## Question 4.5

(3 points)

Assess the null model and hypothesis. Is the drug effective, and how
certain can you be of the results? Can the drug company conclude that
the drug causes a reduction in how likely an individual is to contract
this rare disease?

**Your answer here:** Given the low probability of observing 3 or fewer
disease cases under the null hypothesis, and the fact that this
observation is considered unusually low, we have strong evidence to
reject the null hypothesis. This indicates that the drug is effective in
reducing the likelihood of contracting the disease. Certainty of
Results: The results are statistically significant at the 5% level. We
can be reasonably certain (with a confidence level of at least 95%) that
the drug has an effect. Conclusion: Based on this analysis, the drug
company can conclude that the drug likely causes a reduction in the
likelihood of contracting this rare disease. However, it is important to
consider other factors such as sample size, potential biases, and the
need for further studies to confirm the findings.

# Question 5

(6 points)

Consider a continuous random variable $X$ with the probability density
function defined by $f(x)=\frac{10}{x^2}$ for $x\in [10,\infty)$ and
$f(x)=0$ otherwise. Please give the expected value $E[X]$ of $X$ if
possible, or explain why it is not possible.

**Your answer here:** Given the PDF $f(x) = \frac{10}{x^2}$ for
$x \in [10, \infty)$ and $f(x) = 0$ otherwise, we can find the expected
value $E[X]$ by integrating over the range where $f(x)$ is non-zero:
$E[X] = \int_{10}^{\infty} x \cdot \frac{10}{x^2} \, dx$. Simplify the
integrand: $E[X] = \int_{10}^{\infty} \frac{10}{x} \, dx$. Now,
calculate the integral:
$E[X] = 10 \int_{10}^{\infty} \frac{1}{x} \, dx = 10 \left[ \ln|x| \right]_{10}^{\infty}$.
Evaluate the integral:
$E[X] = 10 \left( \lim_{x \to \infty} \ln|x| - \ln|10| \right)$. Since
$\ln|x|$ approaches infinity as $x$ approaches infinity:
$E[X] = 10 \left( \infty - \ln|10| \right) = \infty$. Therefore, the
expected value $E[X]$ is not finite; it is infinite. This means that the
expected value does not exist in the conventional sense because the
integral diverges.

# Question 6

Consider the scenarios below on estimation, and answer the questions in
each scenario.

## Question 6.1

(5 points)

If you model the data below (represented as b) as the result of 8
independent Bernoulli trials with probability of success equal to $p$,
what is the maximum likelihood estimate of $p$? Please show your work
mathematically.

```{r}
b <- c("success", "success", "failure", "failure", "failure", "success","failure", "failure")
```

**Your answer here:** Given the data $b$ as the result of 8 independent
Bernoulli trials, we want to find the maximum likelihood estimate (MLE)
of the probability of success $p$. Let $X_i$ be the outcome of the
$i^{th}$ trial, where $X_i = 1$ if the trial is a success and $X_i = 0$
if the trial is a failure. The probability mass function for a Bernoulli
random variable is: $P(X_i = x_i) = p^{x_i} (1 - p)^{1 - x_i}$. The
likelihood function for the 8 trials is:
$L(p) = \prod_{i=1}^{8} P(X_i = x_i) = \prod_{i=1}^{8} p^{x_i} (1 - p)^{1 - x_i}$.
Given the data we have:
$\sum_{i=1}^{8} X_i = 3 \quad \text{(number of successes)}$. Thus, the
likelihood function becomes: $L(p) = p^3 (1 - p)^5$. To find the maximum
likelihood estimate of $p$, we take the natural logarithm of the
likelihood function to obtain the log-likelihood function:
$\ell(p) = \ln L(p) = \ln(p^3 (1 - p)^5) = 3 \ln p + 5 \ln(1 - p)$.
Next, we differentiate the log-likelihood function with respect to $p$,
set the derivative to zero to find the critical points, and solve for
$p$: $$
\frac{d\ell(p)}{dp} = \frac{3}{p} - \frac{5}{1 - p} = 0
$$ $$
\frac{3}{p} = \frac{5}{1 - p}
$$ $$
3(1 - p) = 5p
$$ $$
3 - 3p = 5p
$$ $$
3 = 8p
$$ $$
p = \frac{3}{8}
$$ Thus, the maximum likelihood estimate of $p$ is:
$\hat{p} = \frac{3}{8}$.

## Question 6.2

(5 points)

If you model the data below as a sample from a Normal distribution with
mean $\mu = 1.5$, what is the maximum likelihood estimate for the
$\sigma^2$ in the density
$f(x)=\frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{(x-\mu)^2}{2\sigma^2}\right)$
for the data given in the sample? Find the MLE mathematically. Then in
code, use the estimate of $\sigma^2$ to determine the value of the
cumulative distribution $\phi(x)$ such that $\phi(x) = .85$.

```{r}
sample_data = c(-1.1,  3.7,  1.6,  9.2,  1.9,  4.1)
```

**Your answer and code here:**

```{r}
# Given sample data
sample_data = c(-1.1, 3.7, 1.6, 9.2, 1.9, 4.1)
mu = 1.5

# Calculate the MLE for sigma^2
n <- length(sample_data)
sigma2_hat <- (sum((sample_data - mu)^2) / n)
sigma2_hat

# Determine the value of the cumulative distribution phi(x) such that phi(x) = 0.85
phi_x <- qnorm(0.85, mean = mu, sd = sqrt(sigma2_hat))
phi_x
```

# Question 7

The following questions use the data "usa_00022.csv" provided with this
assignment. These data are from The Census Bureau's American Community
Survey (ACS) Public Use Microdata Sample (PUMS) for 2021. They were
downloaded from IPUMS: Steven Ruggles, Sarah Flood, Ronald Goeken, Megan
Schouweiler and Matthew Sobek. IPUMS USA: Version 12.0 [dataset].
Minneapolis, MN: IPUMS, 2022. <https://doi.org/10.18128/D010.V12.0>

The PUMA to county relationships were generated on
<https://mcdc.missouri.edu/applications/geocorr2022.html>

```{r}
dat <- read.csv("usa_00022.csv")
```

## Question 7.1

(3 points)

Please generate a data frame from the full data "usa_00022.csv" which is
restricted to respondents in the Public Use Microsample Areas (PUMAs)
812 through 816 inclusive, that is, cases for which the value of the
variable "PUMA" is greater than or equal to 812 and less than or equal
to 816. These are predominantly in **Denver**, CO. How many cases are in
the restricted data set?

**Your answer and code here:**

```{r}
# Load the necessary library
library(dplyr)

# Filter the data for PUMAs 812 through 816
restricted_data <- dat %>% filter(PUMA >= 812 & PUMA <= 816)

# Get the number of cases in the restricted data set
num_cases <- nrow(restricted_data)
num_cases
```

## Question 7.2

(3 points)

For the data set below of employed individuals not living in group
quarters, please report the slope and intercept of the least squares
best fit line for the model $TRANTIME=m(INCWAGE)+b$, a linear model of
"TRANTIME", top-coded total daily commute time in minutes, on "INCWAGE",
top-coded wage and salary income. The values 999999 and 999998 code
missing data for INCWAGE.

**Your code and answer here**

```{r}
dat.trans <- filter(dat, EMPSTAT == 1, GQ == 1 | GQ == 2, !(INCWAGE %in% c(999999, 999998)))
# Fit the linear model
model <- lm(TRANTIME ~ INCWAGE, data = dat.trans)
# Get the slope and intercept of the model
slope <- coef(model)[2]
intercept <- coef(model)[1]
# Print the results
print(c(intercept, slope))
```

## Question 7.3

(3 points)

Make a scatterplot with the "INCWAGE" variable on the horizontal axis
and the "TRANTIME" variable on the vertical axis. Add the line computed
in above. (If you used the built-in function, please extract the values
of the slope and intercept from the fitted model object, rather than
using copy-paste.) For full credit, please use a strategy to aid
visualization of regions of the plot with many points.

**Your code here:**

```{r}
# Create the scatterplot
ggplot(dat.trans, aes(x = INCWAGE, y = TRANTIME)) +
  geom_point(alpha = 0.25, color = "red") +  # Use transparency to aid visualization
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Add the regression line
  labs(title = "Scatterplot of TRANTIME vs. INCWAGE",
       x = "Income Wage (INCWAGE)",
       y = "Total Daily Commute Time (TRANTIME)") +
  theme_minimal()
```

# Question 8

Data on a quantity can come in the form of the maximum of measurement of
the quantity over multiple observations. To explore this, using the
random seed below, please sample 4 values from the uniform distribution
on [0,1] using "runif". Repeat this 8 times and create a matrix of the
values with the first sample of 4 in the first row, the second in the
second row, and so on. The "byrow" argument to the "matrix" function may
be useful.

## Question 8.1

(3 points)

Please create and print/display the matrix.

**Your code here:**

```{r}
set.seed(12345)
# Sample 4 values from uniform distribution on [0,1], repeated 8 times
samples <- replicate(8, runif(4))
# Create a matrix with the samples, filling by row
sample_matrix <- matrix(samples, nrow = 8, byrow = TRUE)
# Print/display the matrix
print(sample_matrix)
```

## Question 8.2

(4 points)

For the matrix created below, please display a histogram of the maximum
value in each row using the default binwidth.

**Your code here:**

```{r}
set.seed(12345)
mat2 <- matrix(runif(800), ncol = 8)
# Calculate the maximum value in each row
row_max <- apply(mat2, 1, max)
# Display a histogram of the maximum values
hist(row_max, main = "Histogram of Maximum Values in Each Row", 
     xlab = "Maximum Value", ylab = "Frequency", col = "blue")
```

# Question 9

Two discrete random variables $X$ and $Y$ have a joint distribution of
$$
f(x,y)= \frac{x+y+1}{27}
$$ where the values of $x$ can be $0, 1, 2$ and $y$ can be $0, 1, 2$.

*Hint for this question:* For a continuous random variable defined on
sample space $\mathcal{S}$,
$p_X(x) = \int_{y \in \mathcal{S}} p_{X|Y}(x,y) p_Y(y) dy = E_Y[p_{X|Y}(x|y)]$

## Question 9.1

(3 points)

What is $P(X\leq 1, Y = 1)$?

**Your answer here:** Given the joint distribution
$f(x, y) = \frac{x + y + 1}{27}$, we can find the required
probabilities: 1.
$P(X = 0, Y = 1): f(0, 1) = \frac{0 + 1 + 1}{27} = \frac{2}{27}$ 2.
$P(X = 1, Y = 1): f(1, 1) = \frac{1 + 1 + 1}{27} = \frac{3}{27}$ Now, we
sum these probabilities:
$P(X \leq 1, Y = 1) = P(X = 0, Y = 1) + P(X = 1, Y = 1) = \frac{2}{27} + \frac{3}{27} = \frac{5}{27}$.
Therefore, the probability $P(X \leq 1, Y = 1) = \frac{5}{27}$.

## Question 9.2

(3 points)

What is the marginal distribution corresponding to $Y$? I.e. $f(y)$?

**Your answer here:** Given the joint distribution
$f(x, y) = \frac{x + y + 1}{27}$ for $x, y \in \{0, 1, 2\}$, the
marginal distribution $f_Y(y)$ is: $f_Y(y) = \sum_{x=0}^{2} f(x, y)$.
Calculate $f_Y(y)$ for each value of $y$:

1\. For
$y = 0: f_Y(0) = f(0, 0) + f(1, 0) + f(2, 0) = \frac{0 + 0 + 1}{27} + \frac{1 + 0 + 1}{27} + \frac{2 + 0 + 1}{27} = \frac{1}{27} + \frac{2}{27} + \frac{3}{27} = \frac{6}{27} = \frac{2}{9}$

2\. For
$y = 1: f_Y(1) = f(0, 1) + f(1, 1) + f(2, 1) = \frac{0 + 1 + 1}{27} + \frac{1 + 1 + 1}{27} + \frac{2 + 1 + 1}{27} = \frac{2}{27} + \frac{3}{27} + \frac{4}{27} = \frac{9}{27} = \frac{1}{3}$

3\. For
$y = 2: f_Y(2) = f(0, 2) + f(1, 2) + f(2, 2) = \frac{0 + 2 + 1}{27} + \frac{1 + 2 + 1}{27} + \frac{2 + 2 + 1}{27} = \frac{3}{27} + \frac{4}{27} + \frac{5}{27} = \frac{12}{27} = \frac{4}{9}$

Therefore, the marginal distribution of $Y$ is: $$
f_Y(y) = \begin{cases}
\frac{2}{9} & \text{if } y = 0 \\
\frac{1}{3} & \text{if } y = 1 \\
\frac{4}{9} & \text{if } y = 2
\end{cases}
$$

## Question 9.3

(3 points)

What is $P(Y=1)$?

**Your answer here:** From the previous question, we determined the
marginal distribution of $Y$. Therefore, $P(Y=1)$ is given directly by
the marginal distribution of $P(Y=1) = f_Y(1) = \frac{1}{3}$.

## Question 9.4

(3 points)

What is $E(Y)$?

**Your answer here:** From the previous question, we determined the
marginal distribution of $Y$. We can calculate the expected value $E(Y)$
as follows:
$E(Y) = 0 \cdot \frac{2}{9} + 1 \cdot \frac{1}{3} + 2 \cdot \frac{4}{9}$,
then simplify the calculation
$E(Y) = 0 + \frac{1}{3} + \frac{8}{9} = \frac{1}{3} + \frac{8}{9}$ and
convert $\frac{1}{3}$ to a common denominator
$\frac{1}{3} = \frac{3}{9}$. Now, sum the fractions
$E(Y) = \frac{3}{9} + \frac{8}{9} = \frac{11}{9}$. Therefore, the
expected value $E(Y) = \frac{11}{9}$.

## Question 9.5

(3 points)

Are the random variables $X$ and $Y$ independent? Answer by calculating
the marginal distributions of $X$ and $Y$ for full credit.

**Your answer here:** From the previous question, we determined the
marginal distribution of $Y$. To check independence, we need to see if
$f(x, y) = f_X(x) \cdot f_Y(y)$ for all values of $x$ and $y$. For
$(x = 0, y = 0): f(0, 0) = \frac{0 + 0 + 1}{27} = \frac{1}{27}$ and
$f_X(0) \cdot f_Y(0) = \frac{2}{9} \cdot \frac{2}{9} = \frac{4}{81}$.
Clearly, $\frac{4}{81} \neq \frac{1}{27}$. Since
$f_X(0) \cdot f_Y(0) \neq f(0, 0)$, we can conclude that $X$ and $Y$ are
not independent.

# End.
